---
title: "Data Transformation"
author: "Declan Little"
date: "9/18/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 5.1 Introduction

## 5.1.1 Prerequisites

```{r}
library(tidyverse)
library(nycflights13)
```

### 5.1.2

```{r}
flights
```

### 5.1.3 dplyr

-verbs 

filter
arrange
select
mutate
summarize

verb(database,whatToDo?)

# 5.2 Filter Rows with Filter()

Flights that occured in Jan 1st

filter(flights, month==1, day==1)
```{r}
filter(flights, month==1, day==1)
```

```{r}
(jan1 <- filter(flights, month==1, day==1))
```

```{r}
filter(flights, month==12, day==25)
```

```{r}
(santa<-filter(flights, month==12, day==25))
```

##### 5.2.1 Comparisons 

==, >, <, <=, !

```{r}
filter(flights, month==1)
```
USE == instead of =

```{r}
sin(pi/2)
```

```{r}
cos(pi/2)
```
Computer estimates, it doesnt know tht cos(pi/2)=o

```{r}
near(cos(pi/2), 0)
```

##### 5.2.2 Logical Operators

or -----> |
and -----> &
not -----> !

Flights that departed in november **or** december 

**word** is bolded

```{r}
filter(flights, month==11|month==12)
```

may, june, july, august, 

summer<-filter(flights,month==5|month=6|month=7|month=8|)

```{r}
(summer<-filter(flights,month==5|month==6|month==7|month==8))
```

```{r}
filter(flights, month %in% c(5,6,7,8))
```

flights that weren't delayed (on departure or arrival) by more than two hours

filter(flights, dep_delay<120|arr_delay<120)
```{r}
filter(flights, dep_delay<120|arr_delay<120)
```
```{r}
filter(flights, !dep_delay>120|arr_delay>120)
```
Negating it changes it from or to and, and from < to > 

##### 5.2.3 Missing Values

One important feature of R that can make comparisons tricky are missing values, NAs ("not availables") NA represents an unkown value or missing values are "contagious": almost any operation invloving an unknown value will also be unknown. 

```{r}
100==NA
2>NA
x<-NA
is.na(x)
is.na(10)
```
How do I get rid of NA's in my dataset

```{r} 
filter(flights, is.na(dep_time))
filter(flights, !is.na(dep_time))
(withoutNA<-filter(flights, !is.na(dep_time)))
withoutNA

filter(withoutNA, is.na(dep_time))
```

Is empty when you check 

##### 5.2.4 Exercises 

>1. Find all flights that 

1.1Had an arrival delay of two or more hours

```{r}
filter(flights, arr_delay>=120)
```


1.2Flew to Houston (IAH or HOU)

```{r}
filter(flights, dest=="IAH"|dest=="HOU")
filter(flights,dest%in% c("IAH","HOU"))
```


1.3Were operated by United, American, or Delta

```{r}
filter(flights, carrier%in% c("UA","AA","DL"))
```

1.4Departed in summer (July, August, and September)

```{r}
filter(flights, month%in% c(7,8,9))
```


1.5Arrived more than two hours late, but didn’t leave late

```{r}
filter(flights, arr_delay>120&dep_delay<=0)
```

1.6Were delayed by at least an hour, but made up over 30 minutes in flight
Departed between midnight and 6am (inclusive)

ˆ
```{r}
filter(flights, dep_delay>=60, dep_delay - arr_delay>60)
```

Another useful dplyr filtering helper is between(). What does it do? Can you use it to simplify the code needed to answer the previous challenges?
```{r}
filter(flights, between(dep_time, 00, 600))
```


How many flights have a missing dep_time? What other variables are missing? What might these rows represent?

```{r}
filter(flights, is.na(dep_time))
```

Why is NA ^ 0 not missing? Why is NA | TRUE not missing? Why is FALSE & NA not missing? Can you figure out the general rule? (NA * 0 is a tricky counterexample!)


```{r}
flights
```

```{r}
filter(flights, dest=="IAH")
```

##### 5.3 Arrange rows with arrange() 

```{r}
flights
```

```{r}
arrange(flights, dep_delay, month, day)
```
Use desc() to re-order by a column in descending order

```{r}
arrange(flights, desc(dep_delay, month, day))
```

**Missing values are always sorted at the end** 

Template for cleaning database------------>

newDataBase<-filter(database, !is.no(desired_variable))



##### 5.3.1 Ezercises

1.How could you use arrange() to sort all missing values to the start? (Hint: use is.na()).

2.Sort flights to find the most delayed flights. Find the flights that left earliest.

3.Sort flights to find the fastest (highest speed) flights.

speed=distance/time (mph) 

```{r}
arrange(flights, air_time) #incorrect
arrange(flights, desc(distance/air_time))
```


4.Which flights travelled the farthest? Which travelled the shortest?

```{r}
arrange(flights, desc(distance))

arrange(flights, distance)
```
### 3.4 select columns with select()
```{r}
select(flights, year, month, day)
select(flights, year:day)
```
```{r}
select(flights, -(year:day))
```
*-* gives the complement or everything but that variable or column. 

There are a numbe of functions you can use within select (): 

starts_with()
ends_with()
contains()

```{r}
select(flights, starts_with("dep"))
select(flights, ends_with("delay"))
select(flights, contains("day"))

```

```{r}
rename(flights, tail_num=tailnum)
```

```{r}
mydatabase= flights

mydatabase<-rename(flights, tail_num=tailnum)
```


Moving Variables Forward in columns
```{r}
select(flights, dep_delay, arr_delay, everything())
```

1. Brainstorm as many ways as possible to select dep_time, dep_delay, arr_time, and arr_delay from flights.

```{r}
select(flights, starts_with('dep'))
```


2. What happens if you include the name of a variable multiple times in a select() call?


```{r}
select(flights, air_time, distance, air_time)
```

```{r}
flights_narrow <- select(flights, year:day, ends_with("delay"), distance, air_time)

flights_narrow
```

```{r}
mutate(flights_narrow, gain=dep_delay-arr_delay, speed=distance/air_time * 60)

```
You can refer to columns you just created

```{r}
mutate(flights, gain=dep_delay-arr_delay, speed=distance/air_time *60, gain_per_hour=gain/air_time)
```

```{r}
summarize(flights, delay=mean(dep_delay, na.rn=TRUE))
```
```{r}
by_day<-group_by(flights, year, month, day)

```

```{r}
by_carrier<-group_by(flights, carrier)

summarize(by_carrier, delay=mean(dep_delay, na.rm = TRUE))
```


##### 6.1 Combining multiple operations with the pipe 

Imagine that we want to explore the reltionship between distance and average delay for each location. Using what you know about dplyr, you might write code like this. 

Group flights by destination

```{r}
by_dest<-group_by(flights,dest)
```

Summarize the group, using average distance, average delay, and count: 

```{r}
delays<-summarize(by_dest,
                  count=n(),
                  dist=mean(distance, na.rm=TRUE),
                  delay=mean(arr_delay, na.rm=TRUE)
)

delays
                  
```
Visualize your data: 

Make a scatter plot of distance vs. delay, with count mapped to the size aesthetic. 

```{r}
ggplot(data=delays)+ geom_point(mapping=aes(x=dist, y=delay, size=count))
```
Remove noisy points

```{r}
delays<-filter(delays, count>20, dest!="HNL")
```


```{r}
ggplot(data=delays)+ geom_point(mapping=aes(x=dist, y=delay, size=count))
```

Clean up plot

```{r}
ggplot(data=delays, mapping=aes(x=dist, y=delay))+geom_point(aes(size=count), alpha=1/3)+geom_smooth(se=FALSE)
```

```{r pipe}
by_dest<-flights %>% 
  group_by(dest) %>%
  summarize(
    count=n(),
    dist=mean(distance, na.rm=TRUE),
    delay=mean(arr_delay, na.rm=TRUE)
  )%>%
    filter(count>20, dest!="HNL")

ggplot(data=delays, mapping=aes(x=dist, y=delay))+geom_point(aes(size=count), alpha=1/3)+geom_smooth(se=FALSE)
```

##### 5.6.2 Missing Values

You may have wondered about the na.rm argument we used above. What happens if we don't set it?

```{r}
flights %>%
  group_by(year, month, day)%>%
  summarize(delay=mean(dep_delay))
```
is.na(variable) asks a question- is the variable NA?

In this case, where missing values represent cancelled flights, we could also tackle the

```{r}
flights
```

```{r}
not_cancelled<- flights %>%
  filter(is.na(arr_delay),  is.na(dep_delay))

flights
not_cancelled
```
Now, there are some planes that have an average delay of 5 hours (300 minutes)

```{r}
delays<-not_cancelled %>% 
  group_by(tailnum) %>% 
  summarize(delay=mean(arr_delay))

delays
arrange(delays, desc(delay))
```
There are some planes that have an average delay of 300+ minutes
```{r}
delays<-not_cancelled %>%
  group_by(tailnum) %>%
  summarize(delay=mean(arr_delay), 
            count=n()
  )

arrange(delays, desc(delay))
geom_point(mapping=aes(x=count, y=delay), alpha=1/3)
```
```{r}
delays%>%
  filter(count>25)%>%
  ggplot()+
  geom_point(mapping=aes(x=count, y=delay), alpha=.2)
```





